---
title: 'ECON 395: Zillow Part 1'
author: "Brianna Penkala & Kai Hoeger"
date: "2025-01-31"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r, echo = FALSE}
library("dplyr")
library("lubridate")
library("vtable")
library("corrplot")
library("ggplot2")
library("maps")
library("scales")
library("table1")
library("gridExtra")
library("huxtable")
library("rempsyc")
``` 

```{r, echo = FALSE}
# Loading the data
data <- read.csv("/Users/briannapenkala/Downloads/R/Datasets/Zillow Data.xlsx - dataset.csv")

# Converting the date to numeric to make analysis easier
data$date <- as.Date(data$date, format = "%Y.%m.%d")

# Adding a season column
data <- data %>%
  mutate(
    month = month(date), 
    Season = case_when(
      month %in% c(3, 4, 5) ~ "Spring", 
      month %in% c(6, 7, 8) ~ "Summer", 
      month %in% c(9, 10, 11) ~ "Fall", 
      month %in% c(12, 1, 2) ~ "Winter", 
    )
  )
```
<br>
<br>

## 1. Introduction

<br>
<br>
***Data Structure:*** The Zillow data provided is structured in panel form, providing observations about multiple entities over multiple time periods. An observation in this dataset is a region of the country on the last day of each month in a specific year from August 2018 to April 2023 (i.e. Akron, Ohio on August 31, 2018).

***Research Question:*** How do the seasons affect housing prices in the U.S.?

***Outcome Variable:*** The median sales price of houses in the U.S. during the various seasons. 

***Independent Variables:*** As the selling price of a house can be affected by a variety of outside factors other than season, we consider the mean days pending, the median price cut, the number of new listings in the month, the median list price, and the Zillow home value index as additional independent variables to account for. The mean days pending reflects the average number of days pending until a house was sold. A longer pending period suggests more negotiations, possibly leading to the deal falling through, affecting the ultimate sale price. The median price cut refers to the median percentage cut taken on the list price of houses in a particular month. We factor in this variable, as taking a price cut will obviously have implications on the selling price of houses. The number of new listings in the month refers to how many houses were put on the market that month, a factor that would increase competition, and, therefore, influence selling price. We also consider the median list price as, again, this variable would have huge implications on the selling price. The Zillow Home Value Index provides a measure of the “typical home value and market changes across a given region and housing type.” Home value is yet another factor that would affect how much a house is sold for, so it is another important variable to consider.

***Hypothesis:*** We hypothesize that the season with the lowest house prices is the fall. This may be because the weather is getting cold and the houses will have less curb appeal due to the changing seasons. 
<br>
<br>

## 2. Summary/Descriptive Statistics

<br>
<br>
<div align="center">
```{r, echo = FALSE}
# Creating the table

# Selects only numeric data
numeric_data <- data[, -c(1:6, 22:23)]

# Making variable names readable
names(numeric_data) <- c("Median % cut", "Mean % price cut", "Sold in month", "Days pending to sale", "% sold below listing price", "% sold above listing price", "Median sales price to list price", "Median sales price ($)", "Median list price ($)", "Mean days pending", "New pending properties in month", "New listings in month", "Unique listings in month", "Zillow rent index", "Zillow home value index", "Season")

# Refined data for table
table_data <- numeric_data[,c("Median sales price ($)", "Mean days pending", "Median % cut", "New listings in month", "Median list price ($)", "Zillow home value index", "Season")]

# Better stats table https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html
table1(
  ~ . | Season, 
  data = table_data, 
  render.continuous = c("Min", "Max", "Median", "N"), 
  topclass="Rtable1-zebra", 
  render.missing = NULL
)

# would ideally add unit of measurement for those that need it but it isn't given in the data
```
</div>
<br>
<br>

***Summary Statistics Analysis:*** The table reflects the notion that most of the variables are fairly consistent across all months. Overall, home sellers avoid selling in winter, with a sizable drop in the number of new pending properties. Sellers’ avoidance of winter is backed up by the summary statistics as the median percent sold below list price is higher than all of the other months as is the median percent sold above list price. Additionally, the mean days pending is substantially longer - an average of 10 days - as compared to the other seasons. Across the other seasons, summary data of each of the variables fluctuates in terms of which season is “better” for selling a house. General consensus seems to indicate that sellers believe fall and summer are the best months to sell a house as the median number of unique listings are the highest in these two seasons. 
<br>
<br>

## 3. Correlation Matrices

<br>
<br>
<div align = "center">
```{r, echo = FALSE, warning = FALSE}
# Creating correlation matrix (extra parameter to deal with NAs)
corr_matrix <- cor(table_data[1:6], use = "complete.obs")

# Correlation plot (https://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram)
corrplot <- corrplot(corr_matrix, type = "upper", method = "color", order = "hclust", tl.col = "black", tl.srt = 45, addCoef.col = "darkgrey", outline = TRUE)
# (type is layout, hclust is heirarchical ordering)
```
</div>
<br>

***Correlation Matrix Analysis:*** Based on our table, there are a couple of variables that have a strong positive relationship. Median sales price and Zillow home value, median sales price and median list price, and median list price and Zillow home value index are highly correlated to each other. These correlations are expected since the higher median sales or list price will likely result in a higher Zillow home value, as these measurements all represent the value of a home. It is interesting that the mean days pending and median percentage cut are negatively correlated, since this weakly suggests that houses with higher price cuts spend less time pending. It is also interesting that mean days pending is slightly negatively correlated with median sales price and Zillow home value index, suggesting that a higher price may be associated with more days pending. 
<br>
<br>

## 4. Outliers

<br>
<br>
***Methodology:*** In order to identify outliers in our data, we started by creating boxplots of each of our identified variables by season. We decided to split our data seasonally as this is our causal variable of interest and it is necessary to determine if any of our groupings are particularly skewed and would therefore bias our results.

<div align = "center">
```{r, echo = FALSE, warning = FALSE}
# Outlier box plots

# Median % price cut by season
median_cut_box <- ggplot(data, aes(price_cut_perc_median, Season)) + 
  geom_boxplot() + 
  labs(x = "Median % Price Cut")
median_cut_box
```
<br>
```{r, echo = FALSE, warning = FALSE}
# Median list price 
median_list_box <- ggplot(data, aes(median_list_price, Season)) + 
  geom_boxplot() + 
  labs(x = "Median List Price") + 
  scale_x_continuous(labels = comma)
median_list_box
```
<br>
```{r, echo = FALSE, warning = FALSE}
# Mean days pending
mean_pending_box <- ggplot(data, aes(mean_days_pending, Season)) + 
  geom_boxplot() + 
  labs(x = "Mean Days Pending")
mean_pending_box
```
<br>
```{r, echo = FALSE, warning = FALSE}
# Number of new listings
new_listings_box <- ggplot(data, aes(new_listings_count, Season)) + 
  geom_boxplot() + 
  labs(x = "Number of New Listings") + 
  scale_x_continuous(labels = comma)
new_listings_box
```
<br>
```{r, echo = FALSE, warning = FALSE}
# Zillow home value index
zhvi_box <- ggplot(data, aes(zhvi, Season)) + 
  geom_boxplot() + 
  labs(x = "Zillow Home Value Index") + 
  scale_x_continuous(labels = comma)
zhvi_box
```
<br>
```{r, echo = FALSE, warning = FALSE}
# Median sales price 
median_sales_box <- ggplot(data, aes(median_sale_price, Season)) + 
  geom_boxplot() + 
  labs(x = "Median Sales Price") +  
  scale_x_continuous(labels = comma)
median_sales_box
```
</div>
<br>

***Determination:*** As displayed in our graphs, each of the independent variables we identified have numerous outliers. Looking at the tables listing the number of outliers for each season and variable reveals that the outlier count is fairly consistent across the seasons for each of the variables with the exception of increased outliers for the number of new listings in winter and decreased outliers for the price cut taken in summer. All of the outliers that we identified exceeded the upper bound (as opposed to being beneath the lower bound). 


```{r, echo = FALSE}
# Creating a table of outlier totals
outlier_total <- hux(
  "Median Sales Price" = c(79, 72, 79, 83), 
  "Price Cut" = c(24, 22, 9, 15), 
  "List Price" = c(96, 100, 99, 100), 
  "Days Pending" = c(20, 21, 24, 30), 
  "Number of New Listings" = c(150, 134, 120, 134), 
  "ZHVI" = c(84, 75, 74, 86))
```
<br>
<p align="center" style = "font-size: 17px">
<b>Outlier Totals</b>
</p>
```{r, echo = FALSE}
bold(outlier_total)[1,] <- TRUE
bottom_border(outlier_total)[0:1,] <- .4
align(outlier_total) <- "center"
stripe_rows(outlier_total, stripe1 = "white", stripe2 = "grey90")
```
<br>

***Analysis:*** There are a substantial number of outliers in our data, but because the number of outliers are generally evenly distributed across the four seasons throughout each of our explanatory variables, leaving the outliers in our data should not result in substantial bias. Furthermore, our graphs highlight that the ranges of outliers across the seasons for each variable are comparable, providing further evidence that the removal of outliers is not necessary. 
<br>
<br>

## 5. Data Visualization

<br>
<br>
<div align = "center">
```{r, echo = FALSE, warning = FALSE}
# Grouping data by month
month_data <- data %>%
  group_by(month)

# Line graoh of average median sales price by month
line_plot <- ggplot(month_data, aes(month, median_sale_price)) +
  geom_line(stat = "summary", fun = "mean", color = "#466d73") + # calculates the mean of the "value" aka y variable
  scale_x_discrete(limit = c(1:12)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Average Median Sales Cost By Month", 
    x = "Month", 
    y = "Average Cost"
  )

line_plot
```
</div>
<br>


***Line Chart Analysis:*** The chart above highlights the movement of selling price across each of the twelve months of the year. From this visualization, more specific information about the average median selling price of houses can be gleaned. It is interesting to note that the average median sales price plummets from April to May and from July to August. Despite these large declines, spring and summer still lead the way in terms of average median sales price by season as March, April, June, and July report the highest average median selling price data. 

<br>
<br>
<div align = "center">
```{r, echo = FALSE, warning = FALSE}
# Creating averages of sales price for chart   # note using stat = "summary" should automatically do this
sales_avg <- data %>%
  group_by(Season) %>% 
  summarize(avg = mean(median_sale_price))

# Bar chart
season_price_graph <- ggplot(sales_avg, aes(Season, avg)) +
  geom_col(fill = "#9CBABE", width = .5, color = "black", linewidth = .25) +
  scale_y_continuous(labels = comma, limit = c(0, 400000))  + 
  geom_text(aes(label = round(avg, 0)), vjust = -0.5, size = 3) +
  labs(
    title = "Average Median Sales Price by Season",
    x = "Season",
    y = "Average Median Sales Price")

season_price_graph
```
</div>
<br>

***Bar Chart Analysis:*** The bar chart above displays the average median selling price of houses across all years of our data in each season. The visualization provides a quick snapshot of the differences in selling prices of homes by season, highlighting that on average, the selling price of houses in the spring is the highest, followed by summer, winter, and then fall. This is consistent with our hypothesis, as many buyers believe that houses have the most curb appeal in spring, and therefore, will pay more for houses sold during this time period.

```{r, echo = FALSE}
# Loading map data
us_map <- map_data("state")

# Creating lookup table for state abbr to state full name
state_lookup <- data.frame (
  state_abbr = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  region = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", "connecticut", "delaware", "florida", "georgia", 
"hawaii", "idaho", "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", "maine", "maryland", 
"massachusetts", "michigan", "minnesota", "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", "new jersey", 
"new mexico", "new york", "north carolina", "north dakota", "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", "washington", "west virginia", "wisconsin", "wyoming"
))

# Adding full state name column to data (full join adds values even if they aren't matched, therefore NA's for mapping)
map_data <- data %>%
  full_join(state_lookup, by = c("state_name" = "state_abbr"))

```

<div align = "center">
```{r, echo = FALSE}
# Creating averages for each state
state_avg <- map_data[,c(14, 25)] %>%
  group_by(region) %>% 
  summarize(avg = mean(median_sale_price))

# Creating map (use actual map data for map, map_id is our data corresponding to big data, fill is what we want to show)
price_map <- ggplot(state_avg) +
  geom_map(map = us_map, aes(map_id = region, fill = avg), color = "black") +
  labs(
    title = "Average Median Home Sales Price by State",
    x = "Longitude", 
    y = "Latitude", 
    caption = "* White indicates no data was available.") +
  expand_limits(x = us_map$long, y = us_map$lat) +
  scale_fill_continuous(name = "Average price ($)", na.value = "white", low = "#ccf7fc", high = "#466d73", labels = scales::comma_format()) +
  coord_map("mercator") + # prevents the curve distortion of maps
  theme_minimal() 

price_map
```
</div>

***Map Analysis:*** The map above reflects the fact that the selling price of houses is not just dependent on the season, but on the location of the house as well. Darker shaded regions are those where houses sold for higher prices on average, while lighter shaded regions contain houses that sold for lower prices on average. The white states are states that we did not have data for. This visualization is helpful in capturing the effect of outside factors on selling price, displaying that there are likely many influences other than just season on selling price. Using the map visualization also provides useful information for future research projects.
<br>
<br>

## 6. T-tests

<br>
<br>
<p align="center" style = "font-size: 17px">
<b>T-test: Fall and Winter</b>
</p>
```{r, echo = FALSE}
# Isolating two seasons for the first t-test
first_t_data <- data %>%
  filter(Season %in% c("Fall", "Winter"))

# First t-test
first_t <- t.test(median_sale_price ~ Season, data = first_t_data)
first_t <- hux(
  "T-statistic" = -.038267,
  "Degrees of Freedom" = 2817.9, 
  "P-value" = 0.702, 
  "95% Confidence Interval" = "-15119.43 - 10181.67"
)

number_format(first_t)[,2] <- 1

# Table details
bold(first_t)[1,] <- TRUE
bottom_border(first_t)[0:1,] <- .4
align(first_t) <- "center"
stripe_rows(first_t, stripe1 = "white", stripe2 = "grey90")
```
<br>

***T-test Analysis: Fall to Winter:*** The above statistics display the results from a t-test comparing the means of the median selling prices for homes in fall and winter. We chose to compare these two groups because they have the two lowest average median sales prices, as seen in our bar chart. The p-value is very high, 0.702, indicating that we fail to reject the null hypothesis - we cannot be certain that the difference between the means of the median selling prices of homes in fall and winter is different from zero. This suggests that seasonal differences in selling price, at least between fall and winter, are not substantial.
<br>
<br>

<p align="center" style = "font-size: 17px">
<b>T-test: Spring and Summer</b>
</p>
```{r, echo = FALSE}
# Isolating two seasons for the second t-test
second_t_data <- data %>%
  filter(Season %in% c("Spring", "Summer"))

# Second t-test
second_t <- t.test(median_sale_price ~ Season, data = second_t_data)
second_t <- hux(
  "T-statistic" = 0.55396,
  "Degrees of Freedom" = 2522.7, 
  "P-value" = 0.5797, 
  "95% Confidence Interval" = "-9847.142 - 17601.488"
)

number_format(second_t)[,2] <- 1

# Table details 
bold(second_t)[1,] <- TRUE
bottom_border(second_t)[0:1,] <- .4
align(second_t) <- "center"
stripe_rows(second_t, stripe1 = "white", stripe2 = "grey90")
number_format(second_t)[,2] <- 1
```
<br>

***T-test Analysis: Spring to Summer:*** The above statistics display the results from a t-test comparing the means of the median selling prices for homes in spring and summer. We chose to compare these two groups because they have the two highest average median sales prices, as seen in our bar chart. As with the fall to winter comparison, the p-value is high, 0.5797, indicating that we fail to reject the null hypothesis: we cannot be certain that the difference between the means of the median selling prices of homes in spring and summer is different from zero. This p-value is lower than the winter to fall comparison, suggesting that seasonal differences in selling price are more substantial across spring and summer, but are still not statistically significant. 
<br>
<br>

## 7. Conclusion

<br>
<br>
***Summary of Findings:*** Our analysis indicates that in spring, the average selling price of homes is the highest as compared to the other seasons. This is important for sellers to note as they will want to put their homes on the market in spring in order to capture the highest selling price. Specifically, from our line chart, March, April, June, and July appear to be the best months to sell a home in terms of receiving the highest price. We also determined that the mean days pending is the lowest in the winter and the number of new listings is also the lowest in the winter, suggesting less competition. This could suggest to some sellers that they should consider listing their homes in winter, but they would also need to factor in the generally lower average selling price. In future research we would like to consider why selling prices differ seasonally - is our hypothesis regarding better curb appeal in spring the main driver of higher selling prices? What are possible other factors driving seasonal price differences? How come competition appears to have such a small effect on selling price (by average number of listings, winter has the least competition, yet selling prices remain low. Spring has the second least competition, yet selling prices are the highest)? Additionally, we would like to focus on each month - why were there such huge drop offs from month to month (March to April and July to August)?
<br>
<br>
<br>
